Exploring the tree of numerical semigroups

### 2014-10-13 Papier/article.tex
Reference Timing
Read 1 of the paper

\documentclass[reqno,11pt]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{url}
\usepackage{moreverb}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{pgfplots}
\usepackage{alltt}
\usepackage{listings}
\usepackage{xspace}
\usepackage{pdfsync}

@@

@@

@@

A \emph{numerical semigroup} $S$ is a subset of $\NN$ containing $0$, close under addition and of finite complement in $\NN$.  
For example the set 
\begin{equation}
\label{E:NSG}
S_E=\{0,3,6,7,9,10\}\cup\{x\in\NN, x\geq 12\}
\end{equation}
is a numerical semigroup.
The \emph{genus} of a numerical semigroup $S$, denoted by~$g(S)$, is the cardinality of $\NN\setminus S$.
 For example the genus of $S_E$ is $6$,  the cardinality of $\{1,2,4,5,8,11\}$

@@

Here we describe a new algorithm for the exploration of the tree of numerical
semigroup $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$.
The cornerstone of our method is a new combinatorial representation of
numerical semigroup that is well suited for exploration of the
tree~$\mathcal{T}$ and allow large code optimization essentially based on use of
vectorial instructions and parallelization.
The goal of the paper is twofold:
first to present our new encoding of the semigroup and the associated
algorithms, and second to present the optimization techniques which allows, for
those kinds of algorithms, to get speedups by factors of hundreds and even
thousands. We claim that these technique are fairly general for those kinds of
algorithms. As a support for the claim, we applied it to an algorithm of
N.~Borie enumerating integer vector modulo permutation groups~\cite{Borie} and
got a speedup by a factor larger than 2000 using 8 cores !

The paper is divided as follows.
In Section~\ref{S:Tree} we describe the tree of numerical semigroups and give bounds for some parameters attached to a numerical semigroup.
The decription of our new representation of numerical semigroup is done in the third section.
In Section~\ref{S:Algo} we describe an algorithm based on the representation given in Section~\ref{S:DecNumber} and give its complexity. 
Section~\ref{S:Opti} is more technical, and is devoted to the optimisation of the algorithm introduced in Section~\ref{S:Algo}.
Il the last section we emphase the results obtained using our algorithm.

@@

By definition a numerical semigroup is infinite object and we need a finite description of such a object. 
That is the role of generating sets.

@@

$iii)$ Let $x$ be an element of $S\cap\{0,...,c(S)-1\}$. 
The integer $y=c(S)-1-x$ lies in $\{0,...,c(S)-1\}$. 
Moreover $y$ is not in $S$, for otherwise  we write $c(S)-1=y+x$ where $x,y$ are elements of $S$ implying $c(S)-1\in S$. 
In contradiction with the definition of conductor. 
Thus we define an involution 
\[
\begin{array}{rcl}
\psi:\{0,...,c(S)-1\}&\to&\{0,...,c(S)-1\}\\
x&\mapsto&c(S)-1-x
\end{array}
\]
that send $S\{0,...,c(S)-1\}$ into $\{0,...,c(S)-1\}\setminus S$ which is  of cardinality $g(S)$. 
Let us denote by $k$ the cardinality  of $S\cap\{0,...,c(S)-1\}$. 
Since $\psi$ is injective we must have $k\leq g(S)$. 
From the relation 
\[
\{0,...,c(S)-1\}=S\cap\{0,...,c(S)-1\}\sqcup\{0,...,c(S)-1\}\setminus S
\]
 we obtain $c(S)=k+g(S)$ and so $c(S)\leq 2 g(S)$.
\end{proof}

A consequence of Proposition~ \ref{P:Res}~$i)$ is that $\Irr(S)$ is finite and its cardinality is at most $c(S)+m(S)-1$.
More precisely, using Ap\'ery set, we can prove that  the cardinality of $\Irr(S)$ is at most $m(S)$. See Section 2 of Chapter I of~\cite{BookNS} for more details.

@@

\begin{proof}
If $x$ is not irreducible in  $S$, then there exist $a$ and $b$ in $S\setminus{0}$ such that $x=a+b$. 
Since $a\not=0$ and $b\not=0$ hold, the integers  $a$ and $b$ belong to $S\setminus\{x\}$. 
Since $x=a+b$ and $x\not\in S^x$, it follows that  $S^x$ is not stable under addition.

@@

We construct the tree of numerical semigroups, denoted by $\mathcal{T}$ as follows. 
The root of the tree is the unique semigroup of genus $0$, \ie, $\left<1\right>$ that is equal to $\NN$. 
If~$S$ is a semigroup in the tree,  the sons of $S$ are exactly the semigroup~$S^x$ where $x$ belongs to $\Irr(S)\cap[c(s),+\infty[$. 
By convention, when depicting the tree, the numerical semigroup $S^x$ is in the left of $S^y$ if $x$ is smaller than~$y$. 
With this construction, a semigroup $S$ has depth $g$ in $\mathcal{T}$ if and only if its genus is~$g$, see Figure~\ref{F:Tree}.
We denote by $\mathcal{T}_{g}$ the subtree of $\mathcal{T}$ resticted to all semigroup of genus $\leq g$.

\vspace{-3em}

 \begin{figure}[hf!]
\[
\begin{tikzpicture}[level distance=2cm, inner sep=2mm,level/.style={sibling distance=1.55cm}]
    \node {$\sgnode{1}$}
    child {node {$\sgnode{2,3}$}
      child [sibling distance=4.5cm] {node {$\sgnode{3,4,5}$}
        child {node {$\sgnode{4,5,6,7}$}
          child [sibling distance=2cm]{node {$\sgnode{5,6,7,8,9}$}  edge from parent node [left] {4}}
          child [sibling distance=1.9cm]{node {$\sgnode{\gr 4,6,7,9}$}  edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 4,\gr 5,7}$}  edge from parent node [right] {6}}
          child {node {$\sgnode{\gr 4,\gr 5,\gr 6}$}  edge from parent node [right] {7}}
          edge from parent node [left] {3}
        }
        child{{} edge from parent[draw=none]}
        child{{} edge from parent[draw=none]}
        child {node {$\sgnode{\gr 3,5,7}$}
          child {node {$\sgnode{\gr 3,7,8}$} edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 3,\gr 5}$} edge from parent node [right] {7}}
          edge from parent node [left] {4}
        }
        child {node {$\sgnode{\gr 3,\gr 4}$}
          edge from parent node [right] {5}
        }
        edge from parent node [left] {2}
      }
      child [sibling distance=4.5cm] {node {$\sgnode{\gr2,5}$}
        child {node {$\sgnode{\gr2,7}$}
          child {node {$\sgnode{\gr2,9}$}
            edge from parent node [right] {7}
          }
          edge from parent node [right] {5}
        }
        edge from parent node [right] {3}
      }
      edge from parent node [left] {1}
    }
    ;
  \end{tikzpicture}
  \]
\caption{The first four layers of the tree $\mathcal{T}$ of numerical semigroups, corresponding to $\mathcal{T}_4$. A generator of a semigroup is in gray if is not greater that $c(S)$. An edge between a semigroup $S$ and its son $S'$ is labelled by  $x$ is $S'$ is obtained from $S$ by removing~$x$, that is $S'=S^x$.}
\label{F:Tree}
\end{figure}

@@

\begin{exam}
Reconsider the semigroup $S_E$ given at \eqref{E:NSG}.
The integer $14$ admits three decompositions as sums of two elements of $S$, namely $14=0+14$, $14=3+11$ and $14=7+7$. 
Thus the set $D_{S_E}(14)$ is equal to $\{0,3,7\}$ and $d_{S_E}$ equals $3$.
\end{exam}

@@

\begin{proof}
As the set $D_S(x)$ is included in $\{0,...,\left\lfloor\frac x2\right\rfloor\}$, the relation $d_S(x)\leq 1+\left\lfloor \frac{x}2\right\rfloor$ holds.
For $S=\NN$ we have the equality for $D_S(x)$ and so for $d_S(x)$.
\end{proof}

@@

\begin{proof}
We start with $i)$. If $x$ is an element of $S$ then $x$ equals $0+x$. 
The relation~$2\times 0\leq x$ and $0\in S$ imply that $D_S(x)$ contains $0$, and so $d_S(x)>0$ holds.
Conversely, the relation  $d_S(x)>0$  implies that $D_S(x)$ is non-empty. 
Let $y$ be an element of~$D_S(x)$. 
As $y$ and $x-y$ belong to  $S$, by definition, the integer $x=(x-y)+y$ is in $S$  (since $S$ is stable by addition).

Let us show $ii)$. 
Assume $x$ is irreducible in $S$.
There cannot exist $y$ and $z$ in~$(S\setminus\{0\})^2$ such that $x=y+z$. 
The only possible decomposition of $x$ as a sum of two elements of $S$ is $x=0+x$. 
Hence, the set $D_S(x)$ is equal to~$\{0\}$ and we have $d_S(x)=1$. 
Conversely, let $x$ such that $d_S(x)$ is $1$. 
By  $i)$ the  integer $x$ must be in $S$. 
As $x=0+x$ is always a decomposition of $x$ as a sum of two elements in $S$, we obtain $D_S(x)=\{0\}$.
If there exist $y$ and $z$ in $S$ such that $y\leq z$ and $x=y+z$ hold then $y$ lies in $D_S(x)$. 
This implies $y=0$ and $z=x$. 
Hence $x$ is irreducible in~$S$.
\end{proof}

@@

\begin{prop}
Let $S$ be a numerical semigroup and $x$ be an irreducible element of $S$. 
Then for all $y\in \NN\setminus\{0\}$ we have
\[
d_{S^x}(y)=\begin{cases}
d_S(y)-1&\text{if $y\geq x$ and $d_S(y-x)>0$,}\\
d_S(y) &\text{otherwise.}
\end{cases}
\] 
\end{prop}

\begin{proof}
Let $y$ be in $\NN\setminus\{0\}$. We have 
\[
D_{S^x}(y)=\{z \in S^x\ |\ y-z\in S^x\ \text{and}\  2z\leq y\}.
\]
We construct two sets as follow:
\begin{align*}
E&=\{z\in S^x\ |\ y-z=x\ \text{and}\ 2z\leq y\}\\
F&=\begin{cases}
\{x\}&\text{if $y-x\in S$ and $2x\leq y$},\\
\emptyset&\text{otherwise}.
\end{cases}
\end{align*}
We have the relation $D_{S}(y)=D_{S^x}(y)\sqcup E\sqcup F$ and so $d_S(y)$ is equal to $d_{S^x}(y)+\card(E)+\card(F)$
and $D_{S^x}(y)$ is a subset of $D_S(y)$. We start by determining $E$. 
The relation $y-z=x$ implies $z=y-x$ and $2z\leq y$ becomes $y\leq 2x$. Thus we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S^x$ and $y\leq 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
The relation $y-x\in S^x$ together with $y\leq2x$ is equivalent to $y-x\in S$ together with $y<2x$.
Therefore we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S$ and $y< 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
Assume $y-x$ lies in $S$.  If $y<2x$ holds, we have $E=\{y-x\}$ and $F=\emptyset$. Otherwise, we obtain $E=\emptyset$ and $F=\{x\}$. 
Then if $y-x\in S$ holds, we have $\card(E)+\card(F)=1$ and so $d_{S}(y)$ is equal to $d_{S^x}(y)+1$.
If $y-x$ not belong to $S$, then $E$ and $F$ are empty implying $d_{S}(y)=d_{S^x}(y)$. As the condition $y-x\in S$ is equivalent to $y\geq x$ together with $d_S(y-x)>0$, by Proposition~\ref{P:Dandd} the result holds.
\end{proof}

\section{A new algorithm}
\label{S:Algo}
We can easily explore the tree of numerical semigroups up to a genus~$G$ using
a depth first search algorithm (see Algorithm~\ref{A:ExpRec}). This approach does not seem to
have been used before.  In particular, M. Bras-Amor\'os and M.~Delgado use
instead a breadth first search exploration.  The main advantage in our
approach is the small memory needs.  The cost to pay is that, if we want to
explore the tree deeper, we must restart from the root.

@@

Equivalently, we can use an iterative version which uses a stack:
\begin{algorithm}
\caption{\small Iterative Depth first search exploration of the tree $\mathcal{T}_g$}\label{A:Exp}
\small
\begin{algorithmic}[1]
\Procedure{Explore}{\tt{G}}
\State Stack \tt{stack} \Comment{the empty stack}
\State \tt{stack.push($\mathbb{N}$)}
\While{\tt{stack} is not empty} 
	\State $\tt{S} \gets \tt{stack.top()}$
	\State \tt{stack.pop()}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State $\tt{stack.push}(\tt{S}^\tt{x})$
			\EndIf
		\EndFor
	\EndIf
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
By Proposition~\ref{P:Res} $iii)$ we have  the relation $c(S)\leq 2g(S)$ and so the $S$-decomposition number of $c(S)$ occurs in $\delta_S$.
Since $c(S)$ is equal to $\max(\NN\setminus S)$, Proposition~\ref{P:delta} implies 
\[
c(S)=1+\max\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
As all elements of $\NN\setminus S$ are smaller than $c(S)$, their $S$-decomposition numbers are in $\delta_S$ and we obtain
\[
g(S)=\card\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
By Proposition~\ref{P:Res} $ii)$, the relation $m(S)\leq g(S)+1$ holds. 
This implies that the $S$-decomposition number of $m(S)$ appears in $\delta_S$ : 
\[
m(S)=\min\{i\in\{0,...,3G\},\ d_S(i)>0\}.
\]
Since, by Proposition~\ref{P:Res} $i)$,  all irreducible elements are smaller than $c(S)+m(S)-1$, which is itself smaller than $3G$ from the relations $ii)$ and $iii)$ of Proposition~\ref{P:Res}. So Proposition~\ref{P:Dandd} gives
\[
\Irr(S)=\{i \in\{0,...,3G\},\ d_S(i)=1\}.
\]
\end{proof}

@@

For example the following Algorithm initializes a representation of the semigroup $\NN$ ready for an exploration of the tree $\mathcal{T}_G$, \ie, the tree of numerical semigroup upto genus $G$.

\begin{algorithm}
\caption{\small Returns the root of $\mathcal{T}_G$}\label{A:Root}
\begin{algorithmic}
\small
\Function{Root}{\tt{G}}
\State \tt{R.c} $\gets$ 1 \Comment{\tt{R} stands for $\NN$}
\State \tt{R.g} $\gets$ 0
\State \tt{R.m} $\gets$ 1 
\For{\tt{x} from $1$ to $3\,\tt{G}$}
	\State\tt{R.d[x]}$\gets1+\left\lfloor\frac{\tt{x}}2\right\rfloor$
\EndFor
\State{\textbf{return} \tt{R}}
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\newpage

@@

\begin{proof}
Let us check the correctness of the algorithm.
By construction $S^x$ is the semigroup $S\setminus\{x\}$.
Thus the genus $S^x$ is $g(S)+1$, see Line~\tt{1}.
Every integer of $I=[x+1,+\infty[$ lies in $S$ since $x$ is greater than $c(S)$, so the interval $I$ is included in $S^x$.
As $x$ does not belong to $S$, the conductor of $S^x$ is $x+1$, see Line \tt{2}.
For the multiplicity of $S^x$ we have two cases. 
First, if $x>m(S)$ holds  then $m(S)$ is also in $S^x$ and so $m(S^x)$ is equal to $m(S)$.
Assume now $x=m(S)$. 
The relation $x(S)\geq c(S)$ and the characterisation of~$m(S)$ implies $x=m(S)=c(S)$.
Thus $S^x$ contains $m(S)+1$ which is $m(S^x)$.
The initialisation of $m(S^x)$ is done by Lines \tt{4} to \tt{8}. 
The correctness of the computation of $\delta_{S^x}$ (see Proposition~\ref{P:delta}) done from Line \tt{9} to Line \tt{15} is a direct consequence of~Proposition~\ref{P:Dandd}.

@@

For the time complexity, let us remark that Algorithm~\textsc{Son} is called for every semigroup of the tree $\mathcal{T}_G$ (the tree of semigroups of genus $\leq G$). 
Since there are exactly $N=\sum_{g=0}^G n_g$ such semigroups, the time complexity of \textsc{Son} established in Proposition~\ref{P:Son} guarantees that the running time of \textsc{Count} is in~$O(\log(G)\times G\times N)$, as stated.

>>>>

\documentclass[reqno]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{url}
\usepackage{moreverb}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{pgfplots}
\usepackage{alltt}
\usepackage{listings}
\usepackage{xspace}
\usepackage{pdfsync}

@@

\setlength{\oddsidemargin}{0pt} 
\setlength{\evensidemargin}{0pt}
\setlength{\textwidth}{16cm}
\setlength{\textheight}{24.2cm}
\setlength{\topmargin}{0pt}
\setlength{\voffset}{-1.04cm}
\setlength{\headsep}{0.5cm}
\setlength{\headheight}{0pt}

@@

\vspace{1em}

@@

A \emph{numerical semigroup} $S$ is a subset of $\NN$ containing $0$, close under addition and of finite complement in $\NN$.  
For example the set 
\begin{equation}
\label{E:NSG}
S_E=\{0,3,6,7,9,10\}\cup\{x\in\NN, x\geq 12\}
\end{equation}
is a numerical semigroup.
The \emph{genus} of a numerical semigroup $S$, denoted by~$g(S)$, is the cardinality of~$\NN\setminus S$.
 For example the genus of $S_E$ is $6$,  the cardinality of $\{1,2,4,5,8,11\}$

@@

Here we describe a new algorithm for the exploration of the tree of numerical
semigroup $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$.
The cornerstone of our method is a new combinatorial representation of
numerical semigroup that is well suited for exploration of the
tree~$\mathcal{T}$ and allow large code optimization essentially based on use of
vectorial instructions and parallelization.
The goal of the paper is twofold:
first to present our new encoding of the semigroup and the associated
algorithms, and second to present the optimization techniques which allows, for
those kinds of algorithms, to get speedups by factors of hundreds and even
thousands. We claim that these technique are fairly general for those kinds of
algorithms. As a support for the claim, we applied it to an algorithm of
N.~Borie enumerating integer vector modulo permutation groups~\cite{Borie} and
got a speedup by a factor larger than 2000 using 8 cores.

The paper is divided as follows.
In Section~\ref{S:Tree} we describe the tree of numerical semigroups and give bounds for some parameters attached to a numerical semigroup.
The decription of our new representation of numerical semigroup is done in the third section.
In Section~\ref{S:Algo} we describe an algorithm based on the representation given in Section~\ref{S:DecNumber} and give its complexity. 
Section~\ref{S:Opti} is more technical, and is devoted to the optimisation of the algorithm introduced in Section~\ref{S:Algo}.
In the last section we emphase the results obtained using our algorithm.

@@

By definition a numerical semigroup is an infinite object and we need a finite description of such a object. 
That is the role of generating sets.

@@

$iii)$ Let $x$ be an element of $S\cap\{0,...,c(S)-1\}$. 
The integer $y=c(S)-1-x$ lies in $\{0,...,c(S)-1\}$. 
Moreover $y$ is not in $S$, for otherwise  we write $c(S)-1=y+x$ where $x,y$ are elements of $S$ implying $c(S)-1\in S$. 
In contradiction with the definition of conductor. 
Thus we define an involution 
\[
\begin{array}{rcl}
\psi:\{0,...,c(S)-1\}&\to&\{0,...,c(S)-1\}\\
x&\mapsto&c(S)-1-x
\end{array}
\]
that send $S\{0,...,c(S)-1\}$ into $\{0,...,c(S)-1\}\setminus S$ which is  of cardinality~$g(S)$. 
Let us denote by $k$ the cardinality  of $S\cap\{0,...,c(S)-1\}$. 
Since $\psi$ is injective we must have $k\leq g(S)$. 
From the relation 
\[
\{0,...,c(S)-1\}=S\cap\{0,...,c(S)-1\}\sqcup\{0,...,c(S)-1\}\setminus S
\]
 we obtain $c(S)=k+g(S)$ and so $c(S)\leq 2 g(S)$.
\end{proof}

A consequence of Proposition~ \ref{P:Res}~$i)$ is that $\Irr(S)$ is finite and its cardinality is at most $c(S)+m(S)-1$.
More precisely, using Ap\'ery set, we can prove that  the cardinality of $\Irr(S)$ is at most $m(S)$. See Section~2 of Chapter I of~\cite{BookNS} for more details.

@@

\begin{proof}
If $x$ is not irreducible in  $S$, then there exist $a$ and $b$ in $S\setminus{0}$ such that $x=a+b$. 
As $a\not=0$ and $b\not=0$ hold, the integers  $a$ and $b$ belong to $S\setminus\{x\}$. 
Since $x=a+b$ and $x\not\in S^x$, it follows that  $S^x$ is not stable under addition.

@@

We construct the tree of numerical semigroups, denoted by $\mathcal{T}$ as follows. 
The root of the tree is the unique semigroup of genus $0$, \ie, $\left<1\right>$ that is equal to $\NN$. 
If~$S$ is a semigroup in the tree,  the sons of $S$ are exactly the semigroup~$S^x$ where $x$ belongs to $\Irr(S)\cap[c(s),+\infty[$. 
By convention, when depicting the tree, the numerical semigroup $S^x$ is in the left of $S^y$ if $x$ is smaller than~$y$. 
With this construction, a semigroup $S$ has depth $g$ in $\mathcal{T}$ if and only if its genus is~$g$, see Figure~\ref{F:Tree}.
We denote by $\mathcal{T}_{g}$ the subtree of~$\mathcal{T}$ resticted to all semigroup of genus $\leq g$.

 \begin{figure}[hf!]
\[
\begin{tikzpicture}[level distance=2cm, inner sep=2mm,level/.style={sibling distance=1.55cm}]
    \node {$\sgnode{1}$}
    child {node {$\sgnode{2,3}$}
      child [sibling distance=4.5cm] {node {$\sgnode{3,4,5}$}
        child {node {$\sgnode{4,5,6,7}$}
          child [sibling distance=2cm]{node {$\sgnode{5,6,7,8,9}$}  edge from parent node [left] {4}}
          child [sibling distance=1.9cm]{node {$\sgnode{\gr 4,6,7,9}$}  edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 4,\gr 5,7}$}  edge from parent node [right] {6}}
          child {node {$\sgnode{\gr 4,\gr 5,\gr 6}$}  edge from parent node [right] {7}}
          edge from parent node [left] {3}
        }
        child{{} edge from parent[draw=none]}
        child{{} edge from parent[draw=none]}
        child {node {$\sgnode{\gr 3,5,7}$}
          child {node {$\sgnode{\gr 3,7,8}$} edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 3,\gr 5}$} edge from parent node [right] {7}}
          edge from parent node [left] {4}
        }
        child {node {$\sgnode{\gr 3,\gr 4}$}
          edge from parent node [right] {5}
        }
        edge from parent node [left] {2}
      }
      child [sibling distance=4.5cm] {node {$\sgnode{\gr2,5}$}
        child {node {$\sgnode{\gr2,7}$}
          child {node {$\sgnode{\gr2,9}$}
            edge from parent node [right] {7}
          }
          edge from parent node [right] {5}
        }
        edge from parent node [right] {3}
      }
      edge from parent node [left] {1}
    }
    ;
  \end{tikzpicture}
  \]
\caption{The first four layers of the tree $\mathcal{T}$ of numerical semigroups, corresponding to~$\mathcal{T}_4$. A generator of a semigroup is in gray if is not greater than $c(S)$. An edge between a semigroup $S$ and its son $S'$ is labelled by  $x$ is $S'$ is obtained from $S$ by removing~$x$, that is $S'=S^x$.}
\label{F:Tree}
\end{figure}

@@

\begin{exam}
Reconsider the semigroup $S_E$ given at \eqref{E:NSG}.
The integer $14$ admits three decompositions as sums of two elements of $S$, namely $14=0+14$ and $14=7+7$. 
Thus the set $D_{S_E}(14)$ is equal to $\{0,7\}$ and $d_{S_E}$ equals $2$.
\end{exam}

@@

\begin{proof}
As the set $D_S(x)$ is included in $\left\{0,...,\left\lfloor\frac x2\right\rfloor\right\}$, the relation $d_S(x)\leq 1+\left\lfloor \frac{x}2\right\rfloor$ holds.
For $S=\NN$ we have the equality for the set~$D_S(x)$ and so for the integer $d_S(x)$.
\end{proof}

@@

\begin{proof}
We start with $i)$. If $x$ is an element of $S$ then $x$ equals $0+x$. 
The relation~$2\times 0\leq x$ and $0\in S$ imply that $D_S(x)$ contains $0$, and so $d_S(x)>0$ holds.
Conversely, the relation  $d_S(x)>0$  implies that~$D_S(x)$ is non-empty. 
Let $y$ be an element of~$D_S(x)$. 
As $y$ and $x-y$ belong to  $S$, by definition, the integer $x=(x-y)+y$ is in $S$  (since $S$ is stable by addition).

Let us show $ii)$. 
Assume $x$ is irreducible in $S$.
There cannot exist $y$ and $z$ in~$(S\setminus\{0\})^2$ such that~$x=y+z$. 
The only possible decomposition of $x$ as a sum of two elements of $S$ is $x=0+x$. 
Hence, the set $D_S(x)$ is equal to~$\{0\}$ and we have $d_S(x)=1$. 
Conversely, let $x$ such that $d_S(x)$ is $1$. 
By~$i)$ the  integer $x$ must be in $S$. 
As $x=0+x$ is always a decomposition of $x$ as a sum of two elements in $S$, we obtain $D_S(x)=\{0\}$.
If there exist $y$ and $z$ in $S$ such that $y\leq z$ and $x=y+z$ hold then $y$ lies in~$D_S(x)$. 
This implies $y=0$ and $z=x$. 
Hence $x$ is irreducible in~$S$.
\end{proof}

@@

\begin{prop}
Let $S$ be a numerical semigroup and $x$ be an irreducible element of $S$. 
Then for all~$y\in \NN\setminus\{0\}$ we have
\[
d_{S^x}(y)=\begin{cases}
d_S(y)-1&\text{if $y\geq x$ and $d_S(y-x)>0$,}\\
d_S(y) &\text{otherwise.}
\end{cases}
\] 
\end{prop}

\begin{proof}
Let $y$ be in $\NN\setminus\{0\}$. We have 
\[
D_{S^x}(y)=\{z \in S^x\ |\ y-z\in S^x\ \text{and}\  2z\leq y\}.
\]
We construct two sets as follow :
\begin{align*}
E&=\{z\in S^x\ |\ y-z=x\ \text{and}\ 2z\leq y\}\\
F&=\begin{cases}
\{x\}&\text{if $y-x\in S$ and $2x\leq y$},\\
\emptyset&\text{otherwise}.
\end{cases}
\end{align*}
We have the relation $D_{S}(y)=D_{S^x}(y)\sqcup E\sqcup F$ and so $d_S(y)$ is equal to $d_{S^x}(y)+\card(E)+\card(F)$
and $D_{S^x}(y)$ is a subset of $D_S(y)$. We start by determining $E$. 
The relation $y-z=x$ implies $z=y-x$ and $2z\leq y$ becomes $y\leq 2x$. Thus we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S^x$ and $y\leq 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
The relation $y-x\in S^x$ together with $y\leq2x$ is equivalent to $y-x\in S$ together with $y<2x$.
Therefore we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S$ and $y< 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
Assume $y-x$ lies in $S$.  If $y<2x$ holds, we have $E=\{y-x\}$ and $F=\emptyset$. Otherwise, we obtain $E=\emptyset$ and $F=\{x\}$. 
Then if $y-x\in S$ holds, we have $\card(E)+\card(F)=1$ and so $d_{S}(y)$ is equal to $d_{S^x}(y)+1$.
If $y-x$ not belong to $S$, then $E$ and $F$ are empty implying $d_{S}(y)=d_{S^x}(y)$. As the condition $y-x\in S$ is equivalent to $y\geq x$ together with $d_S(y-x)>0$, by Proposition~\ref{P:Dandd} the result holds.
\end{proof}

\section{A new algorithm}
\label{S:Algo}
We can easily explore the tree of numerical semigroups up to a genus~$G$ using
a depth first search algorithm (see Algorithm~\ref{A:ExpRec}). This approach does not seem to
have been used before.  In particular, M.~Bras-Amor\'os and M.~Delgado use
instead a breadth first search exploration.  The main advantage in our
approach is the small memory needs.  The cost to pay is that, if we want to
explore the tree deeper, we must restart from the root.

@@

\noindent Equivalently, we can use an iterative version which uses a stack:

\begin{algorithm}
\caption{\small Iterative Depth first search exploration of the tree $\mathcal{T}_g$}\label{A:Exp}
\small
\begin{algorithmic}[1]
\Procedure{Explore}{\tt{G}}
\State Stack \tt{stack} \Comment{the empty stack}
\State \tt{stack.push($\mathbb{N}$)}
\While{\tt{stack} is not empty} 
	\State $\tt{S} \gets \tt{stack.top()}$
	\State \tt{stack.pop()}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State $\tt{stack.push}(\tt{S}^\tt{x})$
			\EndIf
		\EndFor
	\EndIf
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
By Proposition~\ref{P:Res} $iii)$ we have  the relation $c(S)\leq 2g(S)$ and so the $S$-decomposition number of~$c(S)$ occurs in $\delta_S$.
Since $c(S)$ is equal to $1+\max(\NN\setminus S)$ for $S\not=\NN$, Proposition~\ref{P:delta} implies 
\[
c(S)=1+\max\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
As all elements of $\NN\setminus S$ are smaller than $c(S)$, their $S$-decomposition numbers are in $\delta_S$ and we obtain
\[
g(S)=\card\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
By Proposition~\ref{P:Res} $ii)$, the relation $m(S)\leq g(S)+1$ holds. 
This implies that the $S$-decomposition number of $m(S)$ appears in $\delta_S$ : 
\[
m(S)=\min\{i\in\{0,...,3G\},\ d_S(i)>0\}.
\]
Since, by Proposition~\ref{P:Res} $i)$,  all irreducible elements are smaller than $c(S)+m(S)-1$, which is itself smaller than $3G$ from the relations $ii)$ and $iii)$ of Proposition~\ref{P:Res}. So Proposition~\ref{P:Dandd} gives
\[
\Irr(S)=\{i \in\{0,...,3G\},\ d_S(i)=1\}.
\]
\end{proof}

@@

For example the following Algorithm initializes a representation of the semigroup $\NN$ ready for an exploration of the tree $\mathcal{T}_G$ (the tree of numerical semigroup of genus at most $G$.)

\begin{algorithm}
\caption{\small Returns the root of $\mathcal{T}_G$}\label{A:Root}
\begin{algorithmic}
\small
\Function{Root}{\tt{G}}
\State \tt{R.c} $\gets$ 0 \Comment{\tt{R} stands for $\NN$}
\State \tt{R.g} $\gets$ 0
\State \tt{R.m} $\gets$ 1 
\For{\tt{x} from $0$ to $3\,\tt{G}$}
	\State\tt{R.d[x]}$\gets1+\left\lfloor\frac{\tt{x}}2\right\rfloor$
\EndFor
\State{\textbf{return} \tt{R}}
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

@@

\begin{proof}
Let us check the correctness of the algorithm.
By construction $S^x$ is the semigroup~$S\setminus\{x\}$.
Thus the genus of $S^x$ is $g(S)+1$, see Line~\tt{3}.
Every integer of $I=[x+1,+\infty[$ lies in $S$ since $x$ is greater than~$c(S)$, so the interval $I$ is included in $S^x$.
As $x$ does not belong to $S$, the conductor of $S^x$ is $x+1$, see Line~\tt{2}.
For the multiplicity of $S^x$ we have two cases. 
First, if $x>m(S)$ holds  then $m(S)$ is also in $S^x$ and so $m(S^x)$ is equal to $m(S)$.
Assume now $x=m(S)$. 
The relation $x(S)\geq c(S)$ and the characterization of~$m(S)$ implies $x=m(S)=c(S)$.
Thus $S^x$ contains $m(S)+1$ which is $m(S^x)$.
The initialization of~$m(S^x)$ is done by Lines \tt{4} to \tt{8}. 
The correctness of the computation of $\delta_{S^x}$ (see Proposition~\ref{P:delta} for the definition of $\delta_{S^x}$) done from Line \tt{9} to Line \tt{15} is a direct consequence of~Proposition~\ref{P:Dandd}.

@@

For the time complexity, let us remark that Algorithm~\textsc{Son} is called for every semigroup of the tree~$\mathcal{T}_G$ (the tree of semigroups of genus $\leq G$). 
Since there are exactly $N=\sum_{g=0}^G n_g$ such semigroups, the time complexity of \textsc{Son} established in Proposition~\ref{P:Son} guarantees that the running time of \textsc{Count} is in~$O(\log(G)\times G\times N)$, as stated.
